Using an iterated block particle filter via \pkg{spatPomp}

### 2022-12-20 ms.Rnw
revised ready for recompute

%% customized math macros
%\newcommand\prob[1]{\mathbb{P}\left[{#1}\right]}
\newcommand\prob{\mathbb{P}}
\newcommand\dd[1]{\mathrm{d}{#1}}
\newcommand\given{{\,\vert\,}}
\newcommand\equals{{{\,}={\,}}}
\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
\newcommand\myto{{\;:\;}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
%\newcommand\BirthDeath{\raisebox{-0.3ex}{\scalebox{1.5}{$\cdot$}}}
%\newcommand\dlta[1]{\Delta{#1}}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\argequals{{\,=\,}}
\newcommand\lags{c}
\newcommand\maxlag{\overline{c}}
\newcommand\nlfList{C}
\newcommand\bigO{\mathcal{O}}
%\newcommand\loglik{\ell}
\newcommand\loglik{\lambda}
\newcommand\loglikMC{\MC{\loglik}}
\newcommand\loglikComponent{c}
\newcommand\R{\mathbb{R}}
\newcommand\param{\,;}
\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
\newcommand\MC[1]{#1^{\mbox{\tiny MC}}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{\mathrm{dist}}
\newcommand\transpose{t}

@@

The IBPF algorithm studied by \citet{ning21-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw21arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.

In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.

@@

%%%%%% iiiiiiii %%%%
<<bm-model,echo=T,eval=T>>=
library(spatPomp)
i <- 2
b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
plot(b)
@

@@

<<ibpf-mle-code,eval=F,echo=T>>=
rho_start <- seq(from=0.2,to=0.8,length=U)
params_start <- coef(b)
params_start[paste0("rho",1:U)] <- rho_start
ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
  ibpf(b,params=params_start,
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
@

@@

The MLE may be of less interest than marginal confidence intervals for each unit-specific parameter.
Therefore, we compute a profile likelihood for each $u$. 
We calculate a profile likelihood for each value of $u$, using Monte Carlo adjusted profile methodology \citep{ionides17,ning21}. 
We compare this with an exact likelihood profile constructed by numerical optimization of the log-likelihood evaluated using the Kalman filter.
The IBPF implementation is identical to the search above, except that the profiled parameter is fixed.
For the profile shown in Figure~\ref{fig:ibpf-prof1-plot}, we first evaluate the likelihood using BPF rather than the Kalman filter, to present methodology applicable to non-Gaussian models.
We then check against the likelihood evaluated via the Kalman filter for the IBPF estimates, and the profile computed directly from the Kalman filter.
These reveal a distinct bias in the IBPF/BPF profile, apparently primarily to do with a bias in likelihood evaluation.
The parameter in question describes a dynamic coupling between the units, and it seems that the blocking procedure breaks some of the coupling and thereby infers a higher value of the coupling parameter that used for the simulation.
The bottom panel of Figure~\label{fig:ibpf-prof1-plot} shows that we can also diagnose this effect using the particle filter, on this small example for which the particle filter is tractable.

<<ibpf-prof,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
rho_start <- seq(from=0.2,to=0.8,length=U)
reps <- switch(i,3,10)
param_start_matrix <- matrix(params_start,byrow=TRUE,
  ncol=length(coef(b)),nrow=reps*length(prof_vals),
  dimnames=list(NULL,names(coef(b))))
param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
  ibpf(b,
    params=param_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof1time <- toc-tic
@

<<ibpf-prof-lik,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof1_lik.rda"),seed=868,{
tic <- Sys.time()
prof1_kf_eval <- sapply(ibpf_prof_searches,bm2_kalman_logLik)
ibpf_bpf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
  )
}
ibpf_pf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(pfilter(bm2fit,Np=switch(i,10,1000)))
  )
}
toc <- Sys.time()
})
prof1eval_time <- toc-tic
@

The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.

@@

##### bpf evaluation
p1bpf <- apply(matrix(apply(ibpf_bpf_eval,2,mean),nrow=reps),2,max)
bpf_prof_range <- c(max(p1bpf)-10,max(p1bpf))
plot(y=p1bpf,x=prof_vals,ylim=bpf_prof_range,
  ylab="Log-likelihood")
p1bpf_mcap <- mcap(logLik=p1bpf,parameter=prof_vals)
lines(p1bpf_mcap$fit$parameter,p1bpf_mcap$fit$smoothed)

@@

We now do the same calculation for $\sigma$ for comparison with $\rho$.

<<bm-sigma-model,echo=T,eval=T>>=
set.seed(20)
b_sig <- bm2(U=4,N=switch(i,10,200),unit_specific_names="sigma")

@@

<<ibpf-prof-sig,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof_sig1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
sigma_start <- seq(from=0.4,to=1.6,length=U)
params_sig_start <- coef(b_sig)
params_sig_start[paste0("sigma",1:U)] <- sigma_start
reps <- switch(i,2,10)
param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
  dimnames=list(NULL,names(coef(b_sig))))
param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
  ibpf(b_sig,
    params=param_sig_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="sigma",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof_sig1time <- toc-tic
@

<<ibpf-prof-sig-lik,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof_sig1_lik.rda"),seed=868,{
tic <- Sys.time()
prof_sig1_kf_eval <- sapply(ibpf_prof_sig_searches,bm2_kalman_logLik)
ibpf_bpf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
  )
}
ibpf_pf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(pfilter(bm2fit,Np=switch(i,10,1000)))
  )
}
toc <- Sys.time()
})
prof_sig1eval_time <- toc-tic
@

The $\sigma_1$ profile took \Sexpr{myround(as.double(prof_sig1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof_sig1eval_time,units="mins"),2)}, shared between BPF and PF.

@@

##### bpf evaluation
p1_sig_bpf <- apply(matrix(apply(ibpf_bpf_sig_eval,2,mean),nrow=reps),2,max)
bpf_prof_sig_range <- c(max(p1_sig_bpf)-10,max(p1_sig_bpf))
plot(y=p1_sig_bpf,x=prof_sig_vals,ylim=bpf_prof_sig_range,
  ylab="Log-likelihood")
p1_sig_bpf_mcap <- mcap(logLik=p1_sig_bpf,parameter=prof_sig_vals)
lines(p1_sig_bpf_mcap$fit$parameter,p1_sig_bpf_mcap$fit$smoothed)

@@

These results show that there is noticeable bias in using BPF for likelihood evaluation for this model.
This bias leads to parameter estimation bias for IBPF.
The same methods can be used to assess BPF and IBPF for a metapopulation model.
We will see below that this model is well suited to BPF and IBPF.
Intuitively, this may be because the coupling between units is weak---population movement between towns is critical to disease dynamics, but the vast majority of transmission occurs among residents of the same town.

@@

<<model-he10-code,eval=F,echo=T>>=
he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1964,1964),
  expandedParNames=c("R0"),
  towns_selected=c(1,2,11,12),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=800,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m_seed <- 27
# plot(simulate(he10_model,seed=1))
m <- simulate(he10_model,seed=m_seed)
@

@@

<<m-prof-R0,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_prof_R02.rda"),seed=722,{
m_prof_R0_vals <- seq(from=25,to=35,length=switch(i,5,20))
tic <- Sys.time()
m_R0 <- m
m_R0_prof_u <- 1
reps <- switch(i,3,10)
param_R0_start_matrix <- matrix(coef(m_R0),byrow=TRUE,
  ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
  dimnames=list(NULL,names(coef(m_R0))))
param_R0_start_matrix[,paste0("R0",m_R0_prof_u)] <- rep(m_prof_R0_vals,each=reps)
m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
  ibpf(m_R0,
    params=param_R0_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="R0",
    sharedParNames=NULL,
    block_size=1,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
m_prof_R0_time <- toc-tic
@

<<m-prof-R0-lik,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_prof_R0",m_R0_prof_u,"_lik.rda"),seed=868,{
tic <- Sys.time()
m_bpf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(bpfilter(mfit,block_size=1,Np=switch(i,10,5000)))
  )
}
m_pf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(pfilter(mfit,Np=switch(i,10,5000)))
  )
}
toc <- Sys.time()
})
m_prof_R0_eval_time <- toc-tic
@

The $R0_\Sexpr{m_R0_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_R0_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R0_eval_time,units="mins"),2)}, shared between BPF and PF.

@@

<<model-he10-g,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m-g-sim.rda"),{
he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1964,1964),
  expandedParNames=c("g"),
  towns_selected=c(1,2,11,12),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=800,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m_g <- simulate(he10_model,seed=m_seed)
if(any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
})
@

@@

<<m-slice-g,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_slice_g.rda"),seed=722,{
m_g_prof_u <- 4
slice_reps <- switch(i,2,10)
slice_points <- switch(i,5,20)
slice_g_vals <- rep(seq(from=100,to=2500,length=slice_points),each=slice_reps)
params_g_slice<- matrix(coef(m_g),
  nrow=slice_reps*slice_points,
  ncol=length(coef(m_g)),
  byrow=T,
  dimnames=list(NULL,names(coef(m_g))))
params_g_slice[,paste0("g",m_g_prof_u)] <- slice_g_vals
tic <- Sys.time()
m_slice_g_evals <- foreach(s=1:nrow(params_g_slice),.combine=c)%dopar%{
  logLik(bpfilter(m_g,
    params=params_g_slice[s,],
    Np=switch(i,10,5000),
    block_size=1
  ))
}
toc <- Sys.time()
m_slice_g_time <- toc-tic
})
@

@@

The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.

<<m-prof-g,eval=F,echo=F>>=
stew(file=paste0(m_dir,"m_prof_g.rda"),seed=722,{
m_prof_g_vals <- seq(from=200,to=1800,length=switch(i,5,20))
tic <- Sys.time()
reps <- switch(i,2,10)
param_g_start_matrix <- matrix(coef(m_g),byrow=TRUE,
  ncol=length(coef(m_g)),nrow=reps*length(m_prof_g_vals),
  dimnames=list(NULL,names(coef(m_g))))
param_g_start_matrix[,paste0("g",m_g_prof_u)] <- rep(m_prof_g_vals,each=reps)
m_prof_g_searches <- foreach(s=1:nrow(param_g_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-m_g_prof_u],function(x)paste0("g",x))
  ibpf(m_g,
    params=param_g_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="g",
    sharedParNames=NULL,
    block_size=1,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
m_prof_g_time <- toc-tic
@

<<m-prof-g-lik,eval=F,echo=F>>=
stew(file=paste0(m_dir,"m_prof_g",m_g_prof_u,"_lik.rda"),seed=868,{
tic <- Sys.time()
m_bpf_g_eval <- foreach(mfit=m_prof_g_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(bpfilter(mfit,block_size=1,Np=switch(i,10,5000)))
  )
}
m_pf_g_eval <- foreach(mfit=m_prof_g_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(pfilter(mfit,Np=switch(i,10,5000)))
  )
}
toc <- Sys.time()
})
m_prof_g_eval_time <- toc-tic
@

%% The $g_\Sexpr{m_g_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_g_time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
%% Likelihood evaluation took \Sexpr{myround(as.double(m_prof_g_eval_time,units="mins"),2)}, shared between BPF and PF.

<<m-prof-g-plot,eval=F,echo=F, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $g_4$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit. Bottom: The same IBPF search with likelihood computed using the particle filter.">>=

##### bpf evaluation
m_p2_g_bpf <- apply(matrix(apply(m_bpf_g_eval,2,mean),nrow=reps),2,max)
bpf_prof_g_range <- c(max(m_p2_g_bpf)-10,max(m_p2_g_bpf))
plot(y=m_p2_g_bpf,x=m_prof_g_vals,ylim=bpf_prof_g_range,
  ylab="Log-likelihood",xlab="")
m_p2_g_bpf_mcap <- mcap(logLik=m_p2_g_bpf,parameter=m_prof_g_vals)
lines(m_p2_g_bpf_mcap$fit$parameter,m_p2_g_bpf_mcap$fit$smoothed)

abline(v=m_p2_g_bpf_mcap$ci)

>>>>

%% customized math macros
\newcommand\prob{\mathbb{P}}
\newcommand\dd[1]{\mathrm{d}{#1}}
\newcommand\given{{\,\vert\,}}
\newcommand\equals{{{\,}={\,}}}
\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
\newcommand\myto{{\;:\;}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\argequals{{\,=\,}}
\newcommand\lags{c}
\newcommand\maxlag{\overline{c}}
\newcommand\nlfList{C}
\newcommand\bigO{\mathcal{O}}
\newcommand\loglik{\lambda}
\newcommand\loglikMC{\MC{\loglik}}
\newcommand\loglikComponent{c}
\newcommand\R{\mathbb{R}}
\newcommand\param{\,;}
\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
\newcommand\MC[1]{#1^{\mbox{\tiny MC}}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{\mathrm{dist}}
\newcommand\transpose{t}

@@

The IBPF algorithm studied by \citet{ning21-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw21arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.

@@

%%%%%% iiiiiiii %%%%
<<bm-model,echo=T,eval=T>>=
library(spatPomp)
i <- 1
b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
plot(b)
@

@@

<<ibpf-mle-code,eval=F,echo=T>>=
rho_start <- seq(from=0.2,to=0.8,length=U)
params_start <- coef(b)
params_start[paste0("rho",1:U)] <- rho_start
ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
  ibpf(b,params=params_start,
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
@

@@

The MLE may be of less interest than marginal confidence intervals for each unit-specific parameter.
Therefore, we compute a profile likelihood for each $u$. 
We calculate a profile likelihood for each value of $u$, using Monte Carlo adjusted profile methodology \citep{ionides17,ning21}. 
We compare this with an exact likelihood profile constructed by numerical optimization of the log-likelihood evaluated using the Kalman filter.
The IBPF implementation is identical to the search above, except that the profiled parameter is fixed.
For the profile shown in Figure~\ref{fig:ibpf-prof-rho-plot}, we first evaluate the likelihood using BPF rather than the Kalman filter, to present methodology applicable to non-Gaussian models.
We then check against the likelihood evaluated via the Kalman filter for the IBPF estimates, and the profile computed directly from the Kalman filter.
These reveal a distinct bias in the IBPF/BPF profile, apparently primarily to do with a bias in likelihood evaluation.
The parameter in question describes a dynamic coupling between the units, and it seems that the blocking procedure breaks some of the coupling and thereby infers a higher value of the coupling parameter that used for the simulation.
The bottom panel of Figure~\label{fig:ibpf-prof-rho-plot} shows that we can also diagnose this effect using the particle filter, on this small example for which the particle filter is tractable.

<<ibpf-prof,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
rho_start <- seq(from=0.2,to=0.8,length=U)
reps <- switch(i,3,10)
param_start_matrix <- matrix(params_start,byrow=TRUE,
  ncol=length(coef(b)),nrow=reps*length(prof_vals),
  dimnames=list(NULL,names(coef(b))))
param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
  ibpf(b,
    params=param_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof1time <- toc-tic
@

<<ibpf-prof-lik,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof1_lik.rda"),seed=868,{
tic <- Sys.time()
prof1_kf_eval <- sapply(ibpf_prof_searches,bm2_kalman_logLik)
ibpf_bpf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
  )
}
ibpf_pf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(pfilter(bm2fit,Np=switch(i,10,1000)))
  )
}
toc <- Sys.time()
prof1cores <- cores
})
prof1eval_time <- toc-tic
@

The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{prof1cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.

@@

##### bpf evaluation
p1bpf <- apply(matrix(apply(ibpf_bpf_eval,2,mean),nrow=reps),2,max)
bpf_prof_range <- c(max(p1bpf)-10,max(p1bpf))
plot(y=p1bpf,x=prof_vals,ylim=bpf_prof_range,
  ylab="Log-likelihood",xlab="")
p1bpf_mcap <- mcap(logLik=p1bpf,parameter=prof_vals)
lines(p1bpf_mcap$fit$parameter,p1bpf_mcap$fit$smoothed)

@@

<<bm-sigma-model,echo=F,eval=T>>=
set.seed(20)
b_sig <- bm2(U=4,N=switch(i,10,200),unit_specific_names="sigma")

@@

<<ibpf-prof-sig,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof_sig1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
sigma_start <- seq(from=0.4,to=1.6,length=U)
params_sig_start <- coef(b_sig)
params_sig_start[paste0("sigma",1:U)] <- sigma_start
reps <- switch(i,2,10)
param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
  dimnames=list(NULL,names(coef(b_sig))))
param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
  ibpf(b_sig,
    params=param_sig_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="sigma",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof_sig1time <- toc-tic
@

<<ibpf-prof-sig-lik,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof_sig1_lik.rda"),seed=868,{
tic <- Sys.time()
prof_sig1_kf_eval <- sapply(ibpf_prof_sig_searches,bm2_kalman_logLik)
ibpf_bpf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,10,1000)))
  )
}
ibpf_pf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(pfilter(bm2fit,Np=switch(i,10,1000)))
  )
}
toc <- Sys.time()
prof_sig_cores <- cores
})
prof_sig1eval_time <- toc-tic
@

@@

##### bpf evaluation
p1_sig_bpf <- apply(matrix(apply(ibpf_bpf_sig_eval,2,mean),nrow=reps),2,max)
bpf_prof_sig_range <- c(max(p1_sig_bpf)-10,max(p1_sig_bpf))
plot(y=p1_sig_bpf,x=prof_sig_vals,ylim=bpf_prof_sig_range,
  ylab="Log-likelihood",xlab="")
p1_sig_bpf_mcap <- mcap(logLik=p1_sig_bpf,parameter=prof_sig_vals)
lines(p1_sig_bpf_mcap$fit$parameter,p1_sig_bpf_mcap$fit$smoothed)

@@

We now do the same calculation for $\sigma$, shown in Figure~\ref{fig:ibpf-prof-sig-plot},  for comparison with $\rho$.
For this model, we see that there is noticeable discrepancy between the true profile and that obtained using BPF for likelihood evaluation.
This bias is stronger for $\rho$  than $\sigma$, and may lead to parameter estimation bias for IBPF.
The same methods can be used to assess BPF and IBPF for a metapopulation model.
We will see below that this model is well suited to BPF and IBPF.
Intuitively, this may be because the coupling between units is weak---population movement between towns is critical to disease dynamics, but the vast majority of transmission occurs among residents of the same town.

@@

<<model-he10-code,eval=F,echo=T>>=
he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
  expandedParNames=c("R0"),
  towns_selected=c(1,2,11,12),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=800,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m <- simulate(he10_model,seed=27)
@

@@

<<m-prof-R0,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_prof_R02.rda"),seed=722,{
m_prof_R0_vals <- seq(from=25,to=35,length=switch(i,5,20))
tic <- Sys.time()
m_R0 <- m
m_R0_prof_u <- 1
reps <- switch(i,3,10)
param_R0_start_matrix <- matrix(coef(m_R0),byrow=TRUE,
  ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
  dimnames=list(NULL,names(coef(m_R0))))
param_R0_start_matrix[,paste0("R0",m_R0_prof_u)] <- rep(m_prof_R0_vals,each=reps)
m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
  ibpf(m_R0,
    params=param_R0_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="R0",
    sharedParNames=NULL,
    block_size=1,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
m_prof_R0_time <- toc-tic
@

<<m-prof-R0-lik,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_prof_R0",m_R0_prof_u,"_lik.rda"),seed=868,{
tic <- Sys.time()
m_bpf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(bpfilter(mfit,block_size=1,Np=switch(i,10,5000)))
  )
}
m_pf_R0_eval <- foreach(mfit=m_prof_R0_searches,.combine=cbind) %dopar% {
  replicate(switch(i,2,10),
    logLik(pfilter(mfit,Np=switch(i,10,5000)))
  )
}
toc <- Sys.time()
})
m_prof_R0_eval_time <- toc-tic
prof_R0_cores <- cores
@

The $R0_\Sexpr{m_R0_prof_u}$ profile took \Sexpr{myround(as.double(m_prof_R0_time,units="mins"),2)} mins using \Sexpr{prof_R0_cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(m_prof_R0_eval_time,units="mins"),2)}, shared between BPF and PF.

@@

<<model-he10-g,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m-g-sim.rda"),{
he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
  expandedParNames=c("g"),
  towns_selected=c(1,2,11,12),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=800,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m_g <- simulate(he10_model,seed=27)
if(dim(obs(m_g))!=dim(obs(m_R0)) || any(obs(m_g)-obs(m_R0) != 0)) stop("m_g and m_R0 should have the same synthetic data")
})
@

@@

<<m-slice-g,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_slice_g.rda"),seed=722,{
m_g_prof_u <- 4
slice_reps <- switch(i,2,10)
slice_points <- switch(i,5,20)
slice_g_vals <- rep(seq(from=100,to=2500,length=slice_points),each=slice_reps)
params_g_slice<- matrix(coef(m_g),
  nrow=slice_reps*slice_points,
  ncol=length(coef(m_g)),
  byrow=T,
  dimnames=list(NULL,names(coef(m_g))))
params_g_slice[,paste0("g",m_g_prof_u)] <- slice_g_vals
tic <- Sys.time()
m_slice_g_evals <- foreach(s=1:nrow(params_g_slice),.combine=c)%dopar%{
  logLik(bpfilter(m_g,
    params=params_g_slice[s,],
    Np=switch(i,10,5000),
    block_size=1
  ))
}
toc <- Sys.time()
m_slice_g_time <- toc-tic
m_slice_cores <- cores
})
@

@@

The $g_\Sexpr{m_g_prof_u}$ slice took \Sexpr{myround(as.double(m_slice_g_time,units="mins"),2)} mins using \Sexpr{m_slice_cores} computing cores.
